From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/equivalence.tex
Squeeze some space in equations
Revise sections 6 and 7 and copy-editing

\vc{ The groupoid of finite types is the free symmetric monoidal groupoid on one generator. This can be presented as an
  algebraic 2-theory, which is our syntax for $\PiHatLang$. Vertical categorification of natural numbers as a free
  commutative monoid. See groupoidification. }

@@

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. Since the language have higher-dimensional structure, the syntactic categories here are actually
(locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.

@@

Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig functor from $\PiCat$ to
$\PiPlusCat$.

@@

@@

\todo{Sn/Lehmer/Aut(Fin n) diagram here?}

@@

\begin{definition}[${\gdenot{\blank}} : U \to {\UFin}$]
  \begin{align*}
    \gdenot{\blank}_{0} & = \evalh_{0} \comp \evalp_{0} \comp \evalt_{0} \\
    \gdenot{\blank}_{1} & = \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
  \end{align*}
\end{definition}

@@

\subsection{$\PiLang$ circuit normalisation}

To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote it back to
$\PiPlusLang$. Note that the normalisation happens in going from $\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.

\begin{definition}
  \begin{align*}
    \normt_{0} & : U \to \UPlus                                            \\
    \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0} \\
    \\
    \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
    \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
  \end{align*}
\end{definition}

We now revisit the examples in the introduction but instead of working with circuits operating on sequences of bits, we
work with circuits operating on algebraic data types.

\paragraph*{Normalization by evaluation.} Consider the following reversible function:
$\Afun{RESET}(b,b_1,\ldots,b_n) = (b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is
logical-or and $\underline{\vee}$ is exclusive-or. The function sets or resets the first bit depending on whether any of
the other bits is true or not. The natural definition for this function is recursive: modulo some re-shuffling of the
bits, the strategy is to examine the bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and
terminate; otherwise we continue making recursive calls until we reach the last bit at which point we return the
incoming value with no change:

@@

\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the
introduction. Our Agda infrastructure provides tools to normalize all programs to their normal form. For \Afun{reset 2}
we get:

@@

In principle, the normalized program can be produced following two strategies: (i) by repeatedly applying the rewrite
rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
using a normalization-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
left is compiled to the sequence of four adjacent transpositions on the right:

@@

\paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
permutations. Indeed, instead of writing a program for \Afun{reset 2}, one might simply specify the desired permutation as:

@@

\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
``quoting'' this permutation generates the same normalized program \Afun{reset2Norm} above.

>>>>

@@

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. Since the languages have higher-dimensional structure, the syntactic categories here are actually
(locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.

@@

Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig equivalence from $\PiCat$ to
$\PiPlusCat$.

@@

\begin{proposition}
  $\evalt$ gives a rig biequivalence between $\PiCat$ and $\PiPlusCat$.
\end{proposition}

@@

@@

\begin{definition}[${\gdenot{\blank}} : U \to {\UFin}$]
  \begin{gather*}
    \begin{aligned}
      \gdenot{\blank}_{0} &\defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \gdenot{\blank}_{1} &\defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

@@

\section{Normalisation of reversible circuits}
\label{sec:applications}

Using our semantics, we can normalise a circuit writen in $\PiLang$ -- we first translate it to $\PiHatLang$, then
compute a permutation, and quote it back to $\PiPlusLang$. Note that the normalisation happens in going from
$\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.

\begin{definition}
  \begin{gather*}
    \begin{aligned}
      \normt_{0} & : U \to \UPlus                                            \\
      \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
      \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

We revisit the examples in~\Cref{sec:examples} but instead of working with circuits operating on sequences of bits, we
work with circuits operating on algebraic data types in $\PiLang$.

\paragraph*{Normalisation by evaluation.} Consider the following reversible function: $\Afun{RESET}(b,b_1,\ldots,b_n) =
(b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is logical-or and $\underline{\vee}$ is
exclusive-or. The function sets or resets the first bit depending on whether any of the other bits is true or not. The
natural definition for this function is recursive: modulo some re-shuffling of the bits, the strategy is to examine the
bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and terminate; otherwise we continue making
recursive calls until we reach the last bit at which point we return the incoming value with no change:

@@

\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.
Our Agda infrastructure provides tools to normalise all programs to their normal form. For \Afun{reset 2} we get:

@@

In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
using a normalisation-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
left is compiled to the sequence of four adjacent transpositions on the right:

@@

\paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
as:

@@

\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
``quoting'' this permutation generates the same normalised program \Afun{reset2Norm} above.
